---
title: "Systematic review and climate change analysis in Los Lagos Region"
author: "Derek Corcoran"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    toc: false
bibliography: Biblio.bib
header-includes:
    - \usepackage[round]{natbib}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, cache = T, error = F)

library(ggrepel)
library(patchwork)
library(raster)
library(scales)
library(sf)
library(tidyverse)
library(kableExtra)
```

# Systematic review

```{r}
DF <- read_csv("WebScience_Ref.csv")
colnames(DF) <- make.names(colnames(DF))
```

```{r}
LosLagos <-read_csv("PACL_LL.csv")
```

## Introduction

One of the strategic targets of the Convention of Biological Conservation is "By 2020, at least 17% of terrestrial and inland water, and 10% of coastal and marine areas, especially areas of particular importance for biodiversity and ecosystem services, are conserved through effectively and equitably managed, ecologically representative and well connected systems of protected areas..." 
This emphasizes not only that protected areas (PA) are one of the most effective elements to prevent extinctions hence conserve biological diversity, also future of biological and landscape conservation need planned increment in the number of PA and effectiveness in terms of strategic allocation of management and economic efforts [@le2013protected; @watson2014performance].
This is why the last decades aspects such as economic limitations and the impact on local stakeholders are been taking into account in order to declare news PA and not only its selection based in ecological knowledge [@borrini2004indigenous].

South America has one of the highest proportions of pa area in the world (15.9%, @ProtectedAreas).
In Chile, there are 211 PA in its extension. From those, 187 have national jurisdiction and only 24 international. 
In southern Chile, Los Lagos Region is named given this region has the largest proportion of lake's area in the country, with 52 lakes over 3 Km2, and a total area of	2,850 Km2. It has a population of 828,708 [@Censo2017] and an extension of 48,583.60 Km2.
This region bases its economy in activities related to primary sectors: livestock, aquaculture and forestry industry [@BNC_LosLagos]. Even more, Los Lagos Region presents one of the biggest growth in fisheries and aquaculture growth in comparison with the rest of the country during the last 30 years [@Soto-alvaradoDesarrollo]. 

According to @ProtectedAreas, Los Lagos Region has 21 PA in 15,777 Km2 of surface.
An even when includes a large part of the territory, there has been a disconnection with local communities.
This could be because from those 21 PA in the Region, `r LosLagos %>% dplyr::filter(DESIG %in% c("Monumento Natural", "Parque Nacional")) %>% nrow()` correspond to National Parks (NP) and Natural Monuments (NM) which are IUCN categories that reinforce strict conservation [@dudley2008guidelines]. 
In addition to that, remoteness of PA can generate disconnection. Given most of the PA are located in remote places, people from developed urban centers do not have, in most of the cases, nexus with PA and their territory [@joppa2009high]
Another factor could be a social phenomena, where cultural changes as a product of economic activities such as salmon farming, where the dependency of international markets predominates, transform people in this region do not share a common notion of territory or a regional common agenda (Montecinos, 2009, p. 24)

Among the PA in Los Lagos Region, there are three Andean National Parks: Puyehue National Park (PNP), Vicente Perez Rozales National Park (VPRNP), and Alerce Andino National Park (AANP). PNP, VPRNP, and AANP as all national PA in Chile are under the supervision and administration of the National Forestry Corporation (CONAF). They count with management plans (2008, 2015, and 1997, respectively), which provides a very careful compilation of data. However, these documents have not been correctly updated and they have a lack of information with the changes that have occurred in the last years. For example, between 1995 and 2016 native forests in this area have suffer important losses because of land use changes (mostly land modification to grassland and shrubs, and substitution). 
Hence, the area presents an increment of vulnerability in front of climate change [@marquet2019biodiversidad]. 

Considering relations of PA and local communities, and the importance of the expansion of ecological knowledge to areas that resume the complexity of each PA in order to face specific perturbations or global problematic as climate change in the future, should be a fundamental step in Chile to assure a correct management and projection of PA to the future.
This review revise the ecological literature carry out in these three protected areas and its surroundings in Los Lagos Region in order to generate a base line that allow to project scenarios of climate change to systematic conservation planning in PA.
We made a systematic review of research searching by the names of the three PA: PNP, VPRNP, and AANP; and by the geographical area of Southern Chile.
Doing so, we were able to identified tendencies and gaps in research developed in these three National Parks. We could defined a vision on where to focus new studies that allow systematic conservation plans that includes information to deal with climate change and ensure the biological diversity in Southern Chile.


## Methodology

In order to identify research developed in the focus area, a systematic review of literature was performed through the Web of Knowledge database. Information was categorized according to research areas presented in this project.
Two eligibility criteria were performed: 1. including the name of the National Parks of interest, using keywords at topic level: "Puyehue National Park" or "Vicente Perez Rozales National Park" or "Alerce Andino National Park", and 2. Geographic areas surrounding the National Parks, using keywords at topic level: "South Central Chile" or "North Chilean Patagonia" or "Los Lagos Region". The period of the search included from 1975 until October 2020. Grey-literature was not incorporated in the selection.

Articles selected were categorized by publication year, source title, keywords used, and categories created for the purpose of this review: key component, study location, methodology used, process involved, and study object.
Key components categorized articles according to area of study. This classification included Archaeology & Paleoecology, Climate & environment (including studies in climate, environmental conditions, and seismology), Ecosystem functioning & services, Freshwater biodiversity, Hydrology, Social-ecology, and Species & distribution (including biogeography, species description, and exotic species).

Study location encompassed categories such as the three National Parks of interest: AANP, PNP, and  VPRNP; adjacent areas to NP: Adjacent to AANP, Adjacent to PNP, Adjacent to VPRNP; similar ecosystems to NP:  Andes of Los Rios Region, Argentinian Andes, and Neuquen in Argentina; Larger areas that comprise NP: Los Lagos Region, 2 regions, 3 regions, 4 regions,  5 regions, More than 5 regions; Global studies: Chile, Chile & Argentina; and Close urban centers: Valdivia.

Methodology classified articles based in how data was collected and/or analysed. 
Methods included data from the field: field survey, field collection (similar a survey but with samples that need further processes to be analysed, such as dendrochronology or electric fishing), field experiment (that requires modification to one or more conditions in a location, such as transplant or seedling experiments). 
Other classifications were modeling (including data analysis such as simulations, correlational and multivariate models, and spatial models), molecular analysis (reconstruction of mt-DNA, isotopes, and genetic analysis), and reviews of literature.
There is a category of social methods, that include both quantitative (surveys an questionnaires) and qualitative (interviews and discussion articles).
Process category makes reference to the question behind the articles and implies which is the mechanism studied (see the list of of 47 options in Table  \@ref(tab:tabProcess)).


```{r tabProcess}
Pro <- DF$Process %>% unique() %>% sort() 
Pro <- data.frame(Process = Pro, stringsAsFactors = F)
Pro <- Pro %>% mutate(Process=case_when(Process=="runoff" ~"Runoff",  TRUE ~ Process))
kable(Pro, booktabs = T, caption = "Processes mentioned in studies") 
```


Study object classification was prepared based on what is the minimum element studied in each article. This included living organisms such as humans (household combustion, people, and urban areas), animals (communities of birds, crustacean, fishes, and herpetofauna; species of Aegorhinus, craspedacusta, Dromiciops, Liolaemus, wild boar, and woodwasp), plants (aquatic plants, Chusquea, Embothrium, ferns, native vegetation, pollen, Proteaceae, Sarmienta), other organims (such as cyanobacteria, Didymosphenia,  lichens, and protozoa), and a particular living category given its abundance of publications is forest communities (with focus in one species Austrocedrus, *F. cupressoides*, and Nothofagus; or in more diverse environments as coihue-Rauli-Tepa, forest, native forest, and evergreen trees).

Study objects also included non-living organisms, focusing in geographic elements (basins, groundwater Storage, lakes, runoff, streams, watershed, and also soil and regional subdivision), physical conditions (fires, precipitation, particulate matter, rock art), and data bases.
    
In order to centered this review in ecological knowledge carry out in the area of study, we focus our search in Los Lagos region research and Andean studies in surrounded areas. We left out coastal systems within the region, specially research based on Chiloé island because local communities keep a traditional way of life  and agricultural systems. This was even proposed as Global Importance Agricultural Heritage System (GIAHS) by the Food and Agricultural Organization [@FAO2003Chiloe; @FAO2008Chiloe].


## Results

There is a large amount of research related to Southern Chile. Isi Web of Knowledge database identified 874 records that fitted keywords, of those 520 articles were screened, and  441 full-text articles were assessed for eligibility.

A total of  `r nrow(DF)` articles were selected in this review. From those `r (DF %>% dplyr::filter(Keywords.used=="Name of parks") %>% nrow())` articles were find from the three National Parks and `r (DF %>% dplyr::filter(Keywords.used =="Close area") %>%  nrow())` articles from searching nearby areas of them. 

```{r Years, fig.cap="Temporal distribution of articles according to keywords used in this review of literature", fig.width=8}
Pub_year <-ggplot(DF) + geom_bar(aes(x = Publication.Year, fill=Keywords.used))+  labs(y="Number of articles", x="Publication year")  + theme_classic()+   theme(text=element_text(size=18,  family="Times"))+ scale_fill_manual(values = c('#d8b365','#5ab4ac'))
Pub_year
```


In general terms, number of articles have been increasing during the years. However there are a few years, such as 2003, 2006, 2014, and 2020 that show a striking decrease in publications (Fig \@ref(fig:Years)). Studies in NP do not show a pattern during the years, only two picks in publications: in 2001 and 2015.



```{r Locations, fig.width=8, fig.cap="Locations where studies were carry out"}
DF_Location <- DF %>%  group_by(Study.location) %>% summarise(n = n()) %>% ungroup() %>% dplyr::filter(!is.na(Study.location)) %>% 
  mutate(Study.location=fct_reorder(Study.location, -n))

DF_Location2 <- DF_Location %>% 
  mutate(Study.zone= case_when(Study.location %in% c("PNP", "AANP", "VPRNP")~"National Parks", 
                               Study.location %in% c("Adjacent to PNP", "Adjacent to AANP", "Adjacent to VPRNP")~"Adjacent areas to NP", 
                               Study.location %in% c("Andes of Los Rios Region", "Neuquen, Argentina", "Argentinian andes")~"Similar ecosystems", 
                               Study.location %in% c("Los Lagos Region", "2 regions","3 regions", "4 regions", "5 regions", "More than 5 regions")~"Larger areas that comprise NP", 
                               Study.location %in% c("Chile", "Chile and Argentina")~"Global studies",
                               Study.location %in% c("Valdivia")~"Close urban center")) %>%
  mutate(Study.zone= str_wrap(Study.zone, 10))
                                
Location2 <-ggplot(DF_Location2, aes(x = fct_reorder(Study.zone,n, .desc=TRUE), y = n, group = Study.location)) + geom_col(aes(fill=Study.zone), position="stack") + labs(y="Number of articles", x="Study locations")  + theme_classic()+ ylim(c(0,40))+theme(legend.position = "none")+ scale_fill_manual(values= c('#8c510a','#d8b365','#f6e8c3','#c7eae5','#5ab4ac','#01665e')) + theme(text=element_text(size=18,  family="Times"), axis.text.x = element_text(angle = 0, vjust = 1, hjust=0.5))
Location2
```




From the articles carry out in the NP of interest, only `r (DF %>% dplyr::filter(Study.location == "PNP") %>% nrow())` articles were carry out in PNP, `r (DF %>% dplyr::filter(Study.location == "AANP") %>% nrow())` in AAPN and `r (DF %>% dplyr::filter(Study.location == "VPRNP") %>% nrow())` in VPRPN.
Adjacent areas to these three NP comprehended `r DF %>% dplyr::filter(Study.location %in%  c("Adjacent to PNP", "Adjacent to AANP", "Adjacent to VPRNP")) %>% nrow()` studies. Similar ecosystems, as those found in Andes of Los Ríos Region and Neuquén Province in Argentina, also showed high number of publications `r (DF %>% dplyr::filter(Study.location %in% c("Andes of Los Rios Region", "Neuquen, Argentina", "Argentinian andes")) %>% nrow())`. However, most of the articles (`r (DF %>% dplyr::filter(Study.location %in% c("Los Lagos Region", "2 regions","3 regions", "4 regions", "5 regions", "More than 5 regions")) %>% nrow())`) studied larger areas that comprises Los Lagos Region and the three NP (Fig \@ref(fig:Locations)).



```{r Areas, fig.height=10, fig.width=8, fig.cap="Areas of study. A, study areas for all articles. B, study areas for articles focused in the three NP."}
DF <- DF %>% 
  mutate(Area.of.study2= case_when(Area.of.study %in% c("Arqueological","Paleoecology" )~"Archaeology and Paleoecology",
                                          Area.of.study %in% c("Climate","Environmental conditions", "Seismiology" )~ "Climate and environment",
                                          Area.of.study %in% c("Biogeography","Species", "Exotic species" )~ "Species and distribution",
                                          Area.of.study %in%  "Ecosystem functioning and services" ~ "Ecosystem func. & servic.",
                                          Area.of.study %in% "Freshwater biodiversity" ~ "Freshwater biodiversity",
                                          Area.of.study %in% "Hydrology" ~ "Hydrology",
                                          Area.of.study %in% "Social-ecological" ~ "Social-ecology")) 
DF_Area <- DF%>% 
  mutate(Area.of.study2= str_wrap(Area.of.study2, 15)) %>% 
  group_by(Area.of.study2) %>% summarise(n = n()) %>% ungroup() %>%
   dplyr::filter(!is.na(Area.of.study2))

p1 <-ggplot(DF_Area) + geom_col(aes(x = fct_reorder(Area.of.study2, desc(n)), y=n), fill="#d8b365") +labs(y="Number of articles", x=" ")  + theme_classic()+   theme(text=element_text(size=18,  family="Times"))+ theme(axis.text.x = element_text(angle =90, vjust = 0.5, hjust=1)) 


DF_Area2 <- DF %>% dplyr::filter(Study.location %in% c("PNP", "VPRNP", "AANP")) %>% 
   mutate(Area.of.study2= case_when(Area.of.study %in% c("Arqueological","Paleoecology" )~"Archaeology and Paleoecology",
                                          Area.of.study %in% c("Climate","Environmental conditions", "Seismiology" )~ "Climate and environment",
                                          Area.of.study %in% c("Biogeography","Species", "Exotic species" )~ "Species and distribution",
                                          Area.of.study %in%  "Ecosystem functioning and services" ~ "Ecosystem func. & servic.",
                                          Area.of.study %in% "Freshwater biodiversity" ~ "Freshwater biodiversity",
                                          Area.of.study %in% "Hydrology" ~ "Hydrology",
                                          Area.of.study %in% "Social-ecological" ~ "Social-ecology")) %>% 
  mutate(Area.of.study2= str_wrap(Area.of.study2, 15)) %>% 
  group_by(Area.of.study2, Study.location) %>% summarise(n = n()) %>% ungroup() %>%
   dplyr::filter(!is.na(Area.of.study2))
  

p2 <- ggplot(DF_Area2) + geom_col(aes(x = fct_reorder(Area.of.study2, desc(n)), y=n), fill="#5ab4ac") +labs(y="Number of articles", x=" ")  + theme_classic()+   theme(text=element_text(size=18,  family="Times"))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + facet_wrap(~Study.location, scales="free_x")

p1/p2+ plot_annotation(tag_levels = 'A')
```

According to study areas, half of the studies are focus on ecosystem functioning and services (`r ((((DF %>% dplyr::filter(Area.of.study=="Ecosystem functioning and services") %>% nrow()))/nrow(DF) )*100) %>% round(2)`%). The rest of the studies does not exceed the 20 articles each. They are in decreasing number: Climate and environment, Hydrology, and freshwater biodiversity studies. The rest of the areas have less than 10 articles published (Fig \@ref(fig:Areas)).
In addition, Ecosystem functioning and services has been studied in the three NP, however its proportion in PNP is notably higher than the other NP. Additionally, PNP is the only NP that shows several areas of study, even with only a few articles. In terms of area of study, Hydrology is the next area with more studies in NP with mention in AANP and PNP.



```{r Methods, fig.width=8, fig.cap= "Methods used in selected articles of this review."}
DF <- DF %>% 
  mutate(General_methods= case_when(
    Methodology %in% c("Field collection", "dendrochronology", "Electric fishing")~"Field collection",  
    Methodology %in% c("Field survey", "Sampling stations")~"Field survey", 
    Methodology %in% c("Field experiment", "Transplant experiment", "Seedling experiment", "Stratigraphy")~"Field experiment", 
    Methodology %in% c("Communicational strategy", "Discussion", "Quantitative and qualitative methods", "Survey", "Interview")~"Social methods", 
    Methodology %in% c("Correlations","Modeling", "Multivariate model", "Remote sensing","Simulations", "spatially explicit model")~"Modeling", 
    Methodology %in% c("Chile", "Chile and Argentina")~"Global studies",
    Methodology %in% c("Isotopes","Genetic analysis", "Reconstruction of mtDNA")~"Molecular analysis",
    Methodology %in% c("Review of literature")~"Review of literature"))

DF_Methodology2 <- DF %>%  group_by(General_methods) %>% summarise(n = n()) %>% ungroup() %>% dplyr::filter(!is.na(General_methods)) %>% mutate(General_methods=fct_reorder(General_methods, -n)) %>% mutate(General_methods= str_wrap(General_methods, 10))

p2 <-ggplot(DF_Methodology2, aes(x = fct_reorder(General_methods, n, .desc=TRUE), y = n)) + geom_col(position="dodge", fill="#d8b365") + labs(y="Number of articles", x="Study methods")  + theme_classic()+ ylim(c(0,35))+theme(legend.position = "none") + theme(text=element_text(size=18,  family="Times"), axis.text.x = element_text(angle = 0, vjust = 1, hjust=0.5)) 
  
p2


```

Data collected from the field is abundant and represents a `r (((DF %>% dplyr::filter(General_methods %in% c("Field survey", "Field collection", "Field experiment")) %>% nrow())/nrow(DF))*100) %>% round(2)` % of the studies. Among field collection, it highlights some studies where there is some specific collection of data through methods such as dendrochronology (`r DF %>% dplyr::filter(Methodology =="dendrochronology") %>% nrow()` articles) or electric fishing (`r DF %>% dplyr::filter(Methodology =="Electric fishing") %>% nrow()`), (Fig \@ref(fig:Methods)).
Modeling category included `r DF %>% dplyr::filter(General_methods == "Modeling") %>% nrow()` articles taking into account correlational, multivariate , and spatially explicit models.
Review of literature was mention in `r DF %>% dplyr::filter(General_methods == "Review of literature") %>% nrow()` articles and social methods that encompasses communicational strategies, discussion articles, both quantitative and qualitative methods, surveys, and interviews comprehended only in `r DF %>% dplyr::filter(General_methods == "Social methods") %>% nrow()`. 
The least used method in this review is molecular analysis (only `r DF %>% dplyr::filter(General_methods == "Molecular analysis") %>% nrow()`), and included isotopes, genetic analysis, and reconstruction of mt-DNA.



```{r Objects, fig.width=8, fig.cap= "Objects studied in selected articles of this review."}
DF <- DF %>% 
  mutate(General_object = case_when(
    Study.object %in% c("Household combution","People",  "Urban areas")~"Humans",  
    
    Study.object %in% c("Aegorhinus",  "Bird species", "Craspedacusta", "Crustacean", "Dromiciops","Fishes","Herpetofauna", "Salmonids", "Liolaemus", "Wild boar", "Woodwasp")~"Animals",
    
    Study.object %in% c("Cyanobacteria","Didymosphenia",  "Lichen", "Protozoa")~"Other organims",
    
    Study.object %in% c("Aquatic plants", "Chusquea","Embothrium","Ferns", "Native vegetation", "Pollen", "Proteaceae", "Sarmienta" )~"Plants",
    
    Study.object %in% c("Austrocedrus" ,"F. cupressoides", "Forest","Native forest", "Evergreen trees", "Nothofagus", "Coihue-Rauli-Tepa")~"Forest communities",
    
    Study.object %in% c("Basins", "Groundwater Storage", "Lakes", "runoff", "Soil","Streams", "watershed", "Regional subdivision")~"Geographic elements", 
    
    Study.object %in% c("Fires","Particulate matter","Precipitation", "Rock art")~"Physical conditions",
    Study.object %in% c("Database records", "Spatial proyection", "Temperature records", "Fire database")~"Data base"))

DF_object <- DF %>%  group_by(General_object, Study.object) %>% summarise(n = n()) %>% ungroup()   %>%
    mutate(General_object= str_wrap(General_object, 8)) %>% dplyr::filter(!is.na(Study.object))  

p1 <-ggplot(DF_object, aes(x = fct_reorder(General_object, n, .desc=TRUE), y = n, group = Study.object)) + geom_col(aes(fill=General_object), position="stack") + #geom_text(aes(label =Study.object), size=4, position=position_dodge(0.9), angle=90,vjust=0.5, hjust=-0.04 ) +
labs(y="Number of articles", x="Study object")  + theme_classic()+ ylim(c(0,45))+theme(legend.position = "none")+ scale_fill_brewer(palette = "Dark2") + theme(text=element_text(size=18,  family="Times"), axis.text.x = element_text(angle =0 , vjust = 1, hjust=0.5))+
  scale_fill_manual(values = c('#8c510a','#bf812d','#dfc27d','#f6e8c3','#c7eae5','#80cdc1','#35978f','#01665e'))

p1

```

Living elements from nature are the most frequent object of study in this review (`r ((DF %>% dplyr::filter(General_object%in% c("Animals","Algae, bacterias, and protozoa","Plants", "Forest communities"))%>%  nrow())/nrow(DF)*100) %>% round(2) ` %) (Fig \@ref(fig:Objects)). From those, most of the studies are centered in forest communities (`r DF %>% dplyr::filter(General_object=="Forest communities")%>%  nrow()` articles), where articles that studies Nothofagus (20) and Native forest (15) are the majority (Table \@ref(tab:GeneralObject)).

```{r GeneralObject}
DF1 <- DF %>% 
  mutate(`General object` = case_when(
    Study.object %in% c("Household combution","People",  "Urban areas")~"Humans",  
    
    Study.object %in% c("Aegorhinus",  "Bird species", "Craspedacusta", "Crustacean", "Dromiciops","Fishes","Herpetofauna", "Salmonids", "Liolaemus", "Wild boar", "Woodwasp")~"Animals",
    
    Study.object %in% c("Cyanobacteria","Didymosphenia",  "Lichen", "Protozoa")~"Other organims",
    
    Study.object %in% c("Aquatic plants", "Chusquea","Embothrium","Ferns" , "Native vegetation", "Pollen", "Proteaceae", "Sarmienta" )~"Plants",
    
    Study.object %in% c("Austrocedrus","Coihue-Rauli-Tepa forest" ,"F. cupressoides", "Forest","Native forest", "Evergreen trees", "Nothofagus", "Coihue-Rauli-Tepa")~"Forest communities",
    
    Study.object %in% c("Basins", "Groundwater Storage", "Lakes", "runoff", "Soil","Streams", "watershed", "Regional subdivision")~"Geographic elements", 
    
    Study.object %in% c("Fires","Particulate matter","Precipitation", "Rock art")~"Physical conditions",
    Study.object %in% c("Database records", "Spatial proyection", "Temperature records", "Fire database")~"Data bases"
    ))
  
 DF_objectTable <- DF1 %>%  
  group_by(`General object`, Study.object) %>% summarise(n = n()) %>% ungroup() %>% dplyr::filter(!is.na(Study.object)) %>%
  rename(`Study object`=Study.object, `No of articles`=n) %>% 
  select(`General object`, `Study object`, `No of articles`)


kable(DF_objectTable, booktabs = T, caption = "Study objects in detail", align = "l") %>% 
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")
  

```
Studies in species and communities are mention in three groups: 1. Animals with a `r ((DF %>% dplyr::filter(General_object == "Animals")%>%  nrow())/nrow(DF)*100) %>% round(2) `% of the articles, 2. Plants with a `r ((DF %>% dplyr::filter(General_object == "Plants")%>%  nrow())/nrow(DF)*100) %>% round(2)`%  and 3. Other organisms with a `r ((DF %>% dplyr::filter(General_object == "Animals")%>%  nrow())/nrow(DF)*100) %>% round(2)`% 

Human elements studied only represent `r ((DF %>% dplyr::filter(General_object == "Humans")%>%  nrow())/nrow(DF)*100) %>% round(2) `% of the total of research. This included studies in people (9) and specific interests such as household combustion, and urban areas (only 2 articles)

Geographic objects of study are centered in hydrological elements such as basins, watershed, basins, groundwater storage, runoff, and streams (14 of 16 articles), most of them studied lakes (8 articles).
Physical conditions reunited study objects in a variety from environmental conditions such as precipitation and fire, passing through particular matter and rock art.
  

### Key components

#### Hydrology


```{r data, include=FALSE}
Carpeta_results <- paste0(getwd(), "/Climate_Results/")

### Circulo para GCMComparer

circle <- read_rds(paste0(Carpeta_results,"Circle.rds"))


## Tabla de GCMs

table_scaled <- readRDS(paste0(Carpeta_results,"table_scaled.rds"))

BestNumbers <- function(x, Digits = 2, BigSep = ",", Pre = NULL){
  x <- prettyNum(round(x, digits = Digits), big.mark = BigSep)
  if(!is.null(Pre)){
    x <- paste(Pre, x)
  }
  return(x)
}

AndList <-  function(x, Sep = "and"){
  x <- sort(x)
  First <- paste(x[-length(x)], collapse = ", ")
  Second <- x[length(x)]
  return(paste(First, Second, sep = str_pad(Sep, str_count(Sep) + 2, "both")))
}

Regiones <- read_rds("Regiones.rds")

GCMs <- list.files(path = "ChelsaData/Future") %>% str_remove_all("_rcp85.rds")
```

# Climate change analisis

```{r Crops, cache = T}
### Leer presente

Presente <- read_rds("ChelsaData/Present/Presente.rds") %>% crop(Regiones)

Presente <- Presente  %>% 
  mask(rasterize(Regiones, Presente[[1]]))

Presente <- Presente[[c("Bio1", "Bio12")]]
  
Presente[["Bio1"]] <- Presente[["Bio1"]]/10 

Futuros <- list.files(path = "ChelsaData/Future", full.names = T) %>% purrr::map(read_rds) %>%
purrr::map(~crop(.x, Regiones))

Futuros <- Futuros  %>% purrr::map(~mask(.x, rasterize(Regiones, Presente[[1]])))

Futuros <- Futuros %>% purrr::map(~.x[[c("Bio1", "Bio12")]]) 

for(i in 1:length(Futuros)){
  Futuros[[i]][["Bio1"]] <- Futuros[[i]][["Bio1"]]/10
}

Difs <- Futuros %>% purrr::map(~ .x - Presente)

Dif_Temp <- Difs %>% purrr::map(~.x[["Bio1"]]) %>% reduce(stack)

Dif_Prec <- Difs %>% purrr::map(~.x[["Bio12"]]) %>% reduce(stack)

PNLagos_Rios <- readRDS("~/Documents/CI_Project/PNLagos.rds") %>% dplyr::filter(NAME %in% c("Vicente Perez Rosales", "Puyehue", "Alerce Andino"))

hull <- PNLagos_Rios %>% sf::st_union() %>% st_transform(crs = "+proj=tcea +lon_0=-74.7070313 +datum=WGS84 +units=m +no_defs") %>% st_buffer(10000) %>% st_convex_hull() %>% st_transform("+proj=longlat +datum=WGS84 +no_defs") %>% st_as_sf()
```

## Climate change in southern Chile

In order to evaluate possible scenarios of climate change in the given area, we used a polygon comprising the Los Lagos and Los Ríos regions in Chile, and compared them using GCM compareR [@fajardo2020gcm], considering Mean Anual Temperature and Annual Precipitation. The resulting scaled table of comparisson among futures was then use to select models to be used in the project.

We used the simple structure index (ssi) as implemented by the Vegan package [@Oksanen_2019; @dolnicar1999tale] to test what number of clusters (Between two and eight), was the best way to represent the 32 Compared GCMs. The best representation was five clusters, from each cluster the GCM closest to the centroid of the cluster was selected. The five selected GCMS were `r AndList(table_scaled %>% dplyr::filter(GCM != "") %>% pull(GCM))` and the selected GCMs together with the clusters are shown in figure \@ref(fig:SelectedGCMs). 


```{r SelectedGCMs, fig.cap="In this graph we can see the scaled temperature and precipitation axis, the center of the graph represent the ensemble of all models, the five groups represent the clusters selected using kmeans for five groups, the selected GCM of each cluster is shown with a label"}
ggplot(table_scaled, aes(x = x_axis, y = y_axis)) + 
  geom_point(aes(color = Group)) +
  geom_path(data = circle, aes(x = x, y = y)) + 
  coord_equal() +
  geom_text_repel(aes(label = GCM), nudge_y = 0.3) + 
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  theme_bw() +
  labs(x = "Scaled Temperature Axis",
       y = "Scaled Precipitation Axis")
```

### Present climate conditions

Once the future GCM models were selected, the 30 seconds resolution maps were downloaded for 2070 and for present conditons from CHELSA [@karger2020high].
The present conditions for the Los Lagos and Los Ríos Region are shown in Figure \@ref(fig:PresenteClima), with a close up to the 10 Km buffer surrounding the park in \@ref(fig:PresenteClimaBuffer). The region is a cold and humid area, with a range in the mean annual temperature from `r round(cellStats(Presente[[1]], min), 2)` to `r round(cellStats(Presente[[1]], max), 2)` degrees Celsius and a mean of `r round(cellStats(Presente[[1]], mean), 2)`, and a precipitation range between `r round(cellStats(Presente[[2]], min), 2)` to `r prettyNum(round(cellStats(Presente[[2]], max)),big.mark =",")` and a mean of `r prettyNum(round(cellStats(Presente[[2]], mean)),big.mark = ",")` mm a year. 

```{r PresenteClima, fig.cap= "Mean anual temperature in °C (Facet A), and Annual Precipitation mm (Facet B)", out.width = "100%", out.height="100%"}
Presente_DF <- Presente %>% as("SpatialPixelsDataFrame") %>% as.data.frame()

G1 <- ggplot() + geom_raster(data = Presente_DF, aes(x = x, y = y, fill = Bio1)) + geom_sf(data = Regiones, alpha = 0, size = 0.1) + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1) + geom_sf(data = hull, color = "red", alpha = 0, size = 0.1)+ theme_bw(base_size = 8) + scale_fill_gradient2(name = str_wrap("Mean Annual Temperature [°C]", width = 18), high = muted("red"),
  mid = "white",
  low = muted("blue"),
  midpoint = 0) + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =c(-75:-72))

G2 <- ggplot() + geom_raster(data = Presente_DF, aes(x = x, y = y, fill = Bio12)) + geom_sf(data = Regiones, alpha = 0, size = 0.1) + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1)+ geom_sf(data = hull, color = "red", alpha = 0, size = 0.1) + theme_bw(base_size = 8) + scale_fill_viridis_c(name = str_wrap("Annual Precipitation [mm]", 18)) + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =c(-75:-72)) 

G1 + G2 + plot_annotation(tag_levels = 'A')
```

```{r}
Presente_Hull <- Presente %>% crop(hull) 

Presente_Hull <-  Presente_Hull %>% mask(rasterize(hull, Presente_Hull[[1]]))

Presente_Hull_DF <- Presente_Hull %>%  as("SpatialPixelsDataFrame") %>% as.data.frame()
```


The parks concidered in this proyect are in high altitude which leads to even cooler and wetter conditions, with a range from `r BestNumbers(cellStats(Presente_Hull[[1]], min))` to `r BestNumbers(cellStats(Presente_Hull[[1]], max))` and a mean of `r BestNumbers(cellStats(Presente_Hull[[1]], mean))` degrees Celsius, and a precipitation range from from `r BestNumbers(cellStats(Presente_Hull[[2]], min))` to `r BestNumbers(cellStats(Presente_Hull[[2]], max))` and a mean of `r BestNumbers(cellStats(Presente_Hull[[2]], mean))` mm as seen in figure \@ref(fig:PresenteClimaBuffer).

```{r PresenteClimaBuffer, fig.cap= "Mean anual temperature in °C (Facet A), and Annual Precipitation mm (Facet B) in the three studied national parks white lines, and a 10 km buffer red line", out.width = "100%", out.height="100%"}


G1 <- ggplot() + geom_raster(data = Presente_Hull_DF, aes(x = x, y = y, fill = Bio1)) + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1) + geom_sf(data = hull, color = "red", alpha = 0, size = 0.1)+ theme_bw(base_size = 8) + scale_fill_gradient2(name = str_wrap("Mean Annual Temperature [°C]", width = 18), high = muted("red"),
  mid = "white",
  low = muted("blue"),
  midpoint = 0) + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4))

G2 <- ggplot() + geom_raster(data = Presente_Hull_DF, aes(x = x, y = y, fill = Bio12)) + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1)+ geom_sf(data = hull, color = "red", alpha = 0, size = 0.1) + theme_bw(base_size = 8) + scale_fill_viridis_c(name = str_wrap("Annual Precipitation [mm]", 18)) + labs(x = NULL, y = NULL)+ scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4)) 

G1 + G2 + plot_annotation(tag_levels = 'A')
```

## Future scenarios

### Future temperature

```{r DifTemp, fig.cap= "Changes in future  mean annual temperature for the five selected GCMs, the red polygon surrounds the area of influence while the white line  demarks the limits of the three protected areas in this proyect"}
names(Dif_Temp) <- GCMs
names(Dif_Prec) <- GCMs

Dif_Temp_DF <- Dif_Temp %>% as("SpatialPixelsDataFrame") %>% 
  as.data.frame() %>% 
  pivot_longer(CESM1.BGC:MPI.ESM.LR, names_to = "GCM", values_to = "Temperature")

ggplot() + geom_raster(data =Dif_Temp_DF, aes(x = x, y = y, fill = Temperature)) + geom_sf(data = Regiones, alpha = 0, size = 0.1) + geom_sf(data= hull, alpha = 0, size = 0.1, color = "red") + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1) + facet_wrap(~GCM) + scale_fill_viridis_c(option = "inferno") + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =c(-74:-72)) + theme_bw(base_size = 8)
```

As stated above, the four GCMs chosen to explore and model future scenarios are `r AndList(table_scaled %>% dplyr::filter(GCM != "") %>% pull(GCM))`. Even when this models include relatively wetter models such as `r table_scaled %>% dplyr::filter(GCM != "") %>% dplyr::filter(y_axis == max(y_axis)) %>% pull(GCM)`. On average for the whole region, the temperature will rise from `r BestNumbers(min(cellStats(Dif_Temp, mean)))` to `r BestNumbers(max(cellStats(Dif_Temp, mean)))` depending on the GCM (See figure \@ref(fig:DifTemp)), but in some areas, it the tempearture rise could be as high as `r BestNumbers(max(cellStats(Dif_Temp, max)))` degrees Celsius.

As seen in figure \@ref(fig:DifTempHull), those changes are even higher within the parks and it's sorrounding areas, which means the effects of climate change might be even greater.

```{r DifTempHull, fig.cap="Temperature difference for all five GCMs, for the Close up of the area of influence of the parks"}
Dif_Temp_Hull <- Dif_Temp %>% crop(hull) 

Dif_Temp_Hull <-  Dif_Temp_Hull %>% mask(rasterize(hull, Presente_Hull[[1]]))

Dif_Temp_Hull_DF <- Dif_Temp_Hull %>%  as("SpatialPixelsDataFrame") %>% as.data.frame() %>% pivot_longer(CESM1.BGC:MPI.ESM.LR, names_to = "GCM", values_to = "Temperature")

ggplot() + geom_raster(data =Dif_Temp_Hull_DF, aes(x = x, y = y, fill = Temperature)) + geom_sf(data= hull, alpha = 0, size = 0.1, color = "red") + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1) + facet_wrap(~GCM) + scale_fill_viridis_c(option = "inferno") + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4)) + theme_bw(base_size = 8)
```

### Future precipitation


The change in precipitation is predicted to be much more stark, with some areas decreasing in precipitation up to `r BestNumbers(min(cellStats(Dif_Prec, min)))` as seen in figure \@ref(fig:DifPrec), this is particularly worrisome, since the ecosystems that are prevalent in the area depend on high precipitation.

```{r DifPrec, fig.cap= "Changes in precipitation for the different GCMs, the red polygon surrounds the area of influence while the white line  demarks the limits of the three protected areas in this proyect"}
Dif_Prec_DF <- Dif_Prec %>% as("SpatialPixelsDataFrame") %>% 
  as.data.frame() %>% 
  pivot_longer(CESM1.BGC:MPI.ESM.LR, names_to = "GCM", values_to = "Precipitation")

ggplot() + geom_raster(data =Dif_Prec_DF, aes(x = x, y = y, fill = Precipitation)) + geom_sf(data = Regiones, alpha = 0, size = 0.1) + geom_sf(data= hull, alpha = 0, size = 0.1, color = "red") + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1) + facet_wrap(~GCM) + scale_fill_viridis_c() + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =c(-74:-72)) + theme_bw(base_size = 8)
```





```{r DifPrecHull, fig.cap ="Precipitation difference for all five GCMs, for the Close up of the area of influence of the parks"}

Dif_Prec_Hull <- Dif_Prec %>% crop(hull) 

Dif_Prec_Hull <-  Dif_Prec_Hull %>% mask(rasterize(hull, Presente_Hull[[1]]))

Dif_Prec_Hull_DF <- Dif_Prec_Hull %>%  as("SpatialPixelsDataFrame") %>% as.data.frame() %>% pivot_longer(CESM1.BGC:MPI.ESM.LR, names_to = "GCM", values_to = "Precipitation")

ggplot() + geom_raster(data =Dif_Prec_Hull_DF, aes(x = x, y = y, fill = Precipitation)) + geom_sf(data= hull, alpha = 0, size = 0.1, color = "red") + geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.1) + facet_wrap(~GCM) + scale_fill_viridis_c() + labs(x = NULL, y = NULL) + scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4)) + theme_bw(base_size = 8)
```

The range of changes between GCMs will be from from `r BestNumbers(min(cellStats(Dif_Prec, mean)))` to `r BestNumbers(max(cellStats(Dif_Prec, mean)))` mean annual precipitation for the whole area. Again the areas where there is going to be a higher drop in precipitation are mostly inside of the national parks, as seen in figure \@ref(fig:DifPrecHull), with changes in the mean annual precipitation within the are of `r BestNumbers(min(cellStats(Dif_Prec_Hull, mean)))` for the wettest models, and `r BestNumbers(max(cellStats(Dif_Prec_Hull, mean)))` for the driest models.

## Vegetational formation

```{r}

Formation_Change <- read_rds("Formation_Change.rds")
```


We used @luebert2009depuracion to check current vegetational formations as seen in figure \@ref(fig:VegHull)

```{r VegHull, fig.cap= "Vegetational formation in the influence area of the parks"}
Pisos_hull <- read_rds("Pisos_hull.rds")

Pisos_hull <- Pisos_hull %>% mutate(PISO = str_wrap(PISO, width = 20))

ggplot() + geom_sf(data = Pisos_hull, aes(fill = PISO), size = 0.01) +  geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.5)+ theme(legend.position = "bottom") + scale_fill_viridis_d(name = "Vegetation") + theme_bw(base_size = 8) + scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4))
```

### Climatic analogues

As seen above, changes in precipitation will be very large. Because of that, we checked which areas in the present have the same climate as it is predicted to be in the future in the national parks and sorrounding areas using the analouges R package [@ramirez2011climate]. After doing that, the current vegetational formations of those areas were checked in order to determine possible future vegetation in the area. With that we might better understand the biological consecuences of climate change.
  
```{r}
  NewFormations <- Formation_Change %>% pivot_longer(cols = -c("Rank")) %>% 
    dplyr::filter(!(value %in% Formation_Change$Present)) %>% pull(value) %>% unique()
  
  Desapeared <- Formation_Change %>% pivot_longer(cols = -c("Rank")) %>% 
    dplyr::filter(name != "Present") %>% pull(value) %>% unique()
  
Desapeared <- Formation_Change$Present[!(Formation_Change$Present %in% Desapeared)]
```


Currently in the National Parks and sorrouding areas, the five most common Vegetational formations are `r AndList(Formation_Change$Present)`, and in the future, new vegetational formations that would be part of the five most common ones would include `r AndList(NewFormations)`. Of the current five most common formations, the ones that would be reduced as far as that they would either be lost completely or become very rare are `r AndList(Desapeared)`

```{r TablaForm}
Formation_Change <- read_rds("Formation_Change.rds") %>% mutate_all(~str_wrap(.x, width = 8))

kable(Formation_Change, booktabs = T, caption = "Top five vegetation formations for the present and for the different GCMs acording to the analogous climate analysis") %>% kable_styling(font_size = 5, full_width = F)
```


## Land use

We used @zhao2016detailed landuse layer, specifically developed for Chile. This is a 30 meter resolution, that 

```{r}

Codes <- read_csv("https://raw.githubusercontent.com/derek-corcoran-barrios/UsosDeSueloChile/master/Codes.csv")

LandUse <- raster("/home/derek/Documents/El_Soldado/LC_CHILE_2014_b.tif")

Seldom_Hull <- hull %>% st_transform("+proj=utm +zone=19 +south +datum=WGS84 +units=m +no_defs")

LandUse <- LandUse %>% crop(Seldom_Hull)

LandUse <- LandUse %>% mask(rasterize(Seldom_Hull, LandUse)) 

LandUse2 <- projectRaster(from = LandUse, to = Presente_Hull[[1]], method = "ngb")

LandUse_DF <- LandUse2 %>% as("SpatialPixelsDataFrame") %>% 
  as.data.frame() %>% 
  rename(Code = LC_CHILE_2014_b) %>% 
  left_join(Codes) %>% 
  dplyr::filter(!is.na(Primary_English))


ggplot() + geom_raster(data =LandUse_DF, aes(x = x, y = y, fill =Primary_English)) + scale_fill_viridis_d(name = "Primary landuse") + labs(x = NULL, y = NULL) +
  geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.5) + scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4)) + theme_bw(base_size = 8)
```


```{r}
ggplot() + geom_raster(data =LandUse_DF, aes(x = x, y = y, fill =Secondary_English)) + scale_fill_viridis_d(name = "Secondary landuse") + labs(x = NULL, y = NULL) +
  geom_sf(data = PNLagos_Rios, alpha = 0, color = "white", size = 0.5) + scale_x_continuous(breaks =seq(-72.8, -71.8, by = 0.4)) + theme_bw(base_size = 8)
```


```{r}
LandUse_DF2 <- LandUse_DF %>% group_by(Primary_English) %>%
    summarise(n= n()) %>% 
  mutate(Percentage = round(100*(n/sum(n)),2)) %>% 
  dplyr::select(-n) %>% 
  arrange(desc(Percentage)) %>% 
  rename(`Primary landuse` = Primary_English) %>%
  dplyr::filter(Percentage >= 1)

kable(LandUse_DF2, booktabs = T) %>% kable_styling()
```

```{r}
LandUse_DF3 <- LandUse_DF %>% group_by(Secondary_English) %>%
    summarise(n= n()) %>% 
  mutate(Percentage = round(100*(n/sum(n)),2)) %>% 
  dplyr::select(-n) %>% 
  arrange(desc(Percentage)) %>% 
  rename(`Secondary landuse` = Secondary_English) %>%
  dplyr::filter(Percentage >= 1)

kable(LandUse_DF3, booktabs = T) %>% kable_styling()
```


# References